Donor Responses to Civil Society Repression

### 2025-04-28 index.qmd
Add alt text

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
date: 2025-03-16
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu

@@

```{r}
#| label: fig-agency-response-tree
#| fig-cap: Possible donor- and recipient-country responses to human rights violations
#| include: true
#| fig-width: 6
#| fig-height: 3.8
#| out-width: 80%

@@

Donor countries have a range of possible responses when recipient countries engage in administrative crackdown. Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022a]. @ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25 million after recipient states pass laws that restrict the flow of foreign funds to NGOs. However, aid from multilateral agencies is not affected by these laws. Similarly, @DupuyPrakash:2018 find that the adoption of a restrictive foreign funding law is associated with a 32% decline in aid from bilateral donors, but has no corresponding change in aid from multilateral donors.

@@

Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation, specifically through vaguely defined "political activities" [@Chaudhry:2022]. Donor countries respond differently to these types of oppression. @ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022a]. @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy (i.e. those that fund democracy and civil society promotion activities) reduce funding for advocacy programs by more than 70% in response to new restrictions, and the reduction in aid persists for several years.

@@

```{r}
#| label: fig-individual-response-tree
#| fig-cap: Possible individual responses to civil society restrictions in recipient countries
#| include: true
#| fig-width: 6
#| fig-height: 2.8
#| out-width: 80%

>>>>

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
date: 2025-04-28
author:
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu

@@

```{r}
#| label: fig-agency-response-tree
#| fig-cap: Possible donor- and recipient-country responses to human rights violations
#| fig-alt: |
#|   A flowchart showing possible reactions to human rights oppression.
#|   Recipient governments violate human rights. Donor governments can respond by
#|   doing nothing, threatening the recipient government, withdrawing or suspending
#|   aid, or shifting aid. Recipient governments can counter-respond by doing
#|   nothing, modifying policies to improve relationships, or pressuring donors
#|   to back down.
#| include: true
#| fig-width: 6
#| fig-height: 3.8
#| out-width: 80%

# fig-alt-short: "A flowchart showing possible reactions to oppression: recipient governments violate human rights, donors respond, and recipients counter-respond."

@@

Donor countries have a range of possible responses when recipient countries engage in administrative crackdown. Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022]. @ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25 million after recipient states pass laws that restrict the flow of foreign funds to NGOs. However, aid from multilateral agencies is not affected by these laws. Similarly, @DupuyPrakash:2018 find that the adoption of a restrictive foreign funding law is associated with a 32% decline in aid from bilateral donors, but has no corresponding change in aid from multilateral donors.

@@

Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation, specifically through vaguely defined "political activities" [@Chaudhry:2022]. Donor countries respond differently to these types of oppression. @ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022]. @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy (i.e. those that fund democracy and civil society promotion activities) reduce funding for advocacy programs by more than 70% in response to new restrictions, and the reduction in aid persists for several years.

@@

```{r}
#| label: fig-individual-response-tree
#| fig-cap: Possible individual responses to civil society restrictions in recipient countries
#| fig-alt: |
#|   A flowchart showing possible individual and NGO reactions to civil society
#|   restrictions. NGO host governments restrict civil society. Individual private
#|   donors can respond by doing nothing, increasing their donations, or decreasing 
#|   their donations. NGOs then respond in an unknown way.
#| include: true
#| fig-width: 6
#| fig-height: 2.8
#| out-width: 80%

# fig-alt-short: "A flowchart showing possible individual and NGO reactions to civil society restrictions: governments limit civil society, individual donors respond."
